A primer to numerical simulations: The perihelion motion of Mercury

### 2017-11-30 tex/MercuryNumerical.tex
Inka+ck edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}

@@

\usepackage{ amssymb }
\usepackage{graphicx}           % Import Pictures + PDFs
\usepackage[svgnames]{xcolor}   % Enabling colors by their 'svgnames'
\usepackage{ulem}
\definecolor{fzjblue}{HTML}{005B81}
\definecolor{light-gray}{gray}{0.97}
\definecolor{semilight-gray}{gray}{0.67}
\definecolor{mygreen}{rgb}{0,0.6,0}
\usepackage{listings} %  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings
\lstset{
  language=Python,
  backgroundcolor=\color{light-gray},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\linespread{1}\footnotesize\ttfamily,        % the size of the fonts that are used for the code
  commentstyle=\color{mygreen},
  frame=single,
  keywordstyle=\color{blue},       % keyword style
  tabsize=4,	                   % sets default tabsize to 2 spaces
  showtabs=true,
  rulecolor=\color{black},
  columns=fullflexible,
  literate={\ }{{\copyablespace}}1
}

\usepackage[prependcaption, textsize=tiny]{todonotes}
\usepackage{gensymb}

\newcommand{\python}[0]{\texttt{Python} }
\newcommand{\abs}[1]{\left\vert #1 \right\vert}

\begin{document}

@@

\author{I. Hammer, C. Hanhart, C. K\"orber and C. M\"uller}

\address{Forschungszentrum J\"ulich}
\ead{c.hanhart@fz-juelich.de}
\vspace{10pt}
%\begin{indented}
%\item[]February 2014
%\end{indented}

@@

@@

We are convinced that in order to excite the students for numerical simulations it is compulsory to
demonstrate their power on an example that catches their interest. This purpose is served 
perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to
a very broad public. While their detailed study needs deeper knowledge in Differential Geometry, for
the study outlined here very little math is necessary, such that high school students from about 
10$^{\rm th}$ grade up should benefit from it. This was already demonstrated in the
'Sch\"ulderakademie Teilchenphysik', where an early version of this course was tested successfully 
on a goup of 10 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.

@@

We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories
observed. In other words, we need to show that we can calculate the location in space of the object of interest at any point
in time, once the initial conditions are fixed properly. If we can neglect the finite size of this object (and
in particular its orientation in space) the location is parametrized
by a single three vector $\vec x(t)$. As will become clear below, in order to describe and control the dynamics of a physical
object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}\label{eq:x_update}
\vec x(t+\Delta t) = \vec x(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with
higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec x(t)}{\Delta t} =: \frac{d\vec x(t)}{dt}  = \dot{\vec  x}(t) \ ,
\end{equation}
where $\Delta \vec x(t)=\vec x(t+\Delta t)-\vec x(t)$ and we introduced with the last expression a common short hand notation
for time derivatives. Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec x(t)}{dt^2}  = \ddot{\vec x}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a 
constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law. 
Said differently: a force $\vec F$ shows up by changing the motion of some object. This is quantified in Newton's second law
\begin{equation}
\vec F(\vec x, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation} 
If the mass does not change~\footnote{A well known example where $m$ does change with time is a 
rocket, whose mass decreases as the rocket rises.} with time, this reduces to the well known
\begin{equation}
\vec F(\vec x) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{x}}(t) \ . \label{eq:newton2}
\end{equation} 
Note that in general the force could depend also on the time or the velocity. We here restrict ourselves to the
case relevant for our example, where the force depends on the location only. 
Therefore, 
as soon as the force $\vec F(\vec x)$ is known for all $\vec x$ one can in principle calculate the trajectory, e.g. solving eq.~(\ref{eq:newton2})
for $\vec x(t)$.
Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists. However, alternatively 
one can calculate the whole trajectory
of some test body that experiences this force by a successive application of the rules 
given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
%
\todo{See previous comment.}
%
\begin{enumerate}
\item For a given time $t$, where $\vec x(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec x(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec x(t_0)$ as well as $\vec v(t_0)$ must be known --- the 
trajectories depend on these initial conditions.~\footnote{In general a differential equation of $n^{\rm th}$ degree (where
the highest derivative is $n$) needs $n$ initial conditions specified. For $n=2$ those are often chosen as location and velocity
at some defined time, but one may as well pick two location at different times. }

For this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
studied. One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec x(t)) |\Delta t\ . 
\label{eq:check}
\end{equation}
holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.  Eq.~({\ref{eq:check}})
also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since
the time steps need to be small enough that all changes induced by the force get resolved. Clearly,
a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of
the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing
$\Delta t$ by $\Delta t/2$ should not change the result significantly.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For this concrete example the starting point for the force is Newtons law
of gravitation
\begin{equation}
\vec F_N(\vec x) = - \frac{G_N m M_\odot}{r^2} \frac{\vec x}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation,
$m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec x(t)|$ denotes the distance between sun and Mercury, when we assume that the sun
is infinitely more heavy than the planet and located at the center of the coordinate system. 
Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
%
\todo{This possibly can become a problem for our accuracy because it is at the 10\% level of our expected result, right? Same for $c$.}
%
For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the sun
for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and
$c$. This is what we need $r_S$ for later in the discussion. That the Schwarzschild radius is
also an important quantity to characterize black holes is not relevant for the discussion
at hand.
With this Newtons second law reads 
\todo{Or would it be better to write: $\ddot{x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)$ and $\ddot{\vec x} = - \ddot{x} \frac{\vec{x}}{r}$?}
\begin{equation}
\ddot{\vec x} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \frac{\vec{x}}{r} \ .
\label{eq:newton}
\end{equation}
%\begin{itemize}
%\item{\bf First task:}
%Now the students should implement the program outlined in Sec.~\ref{sec:tva} and visualize the resulting
%trajectories of Mercury for different start parameters.
%Why is it sufficient to work in two dimensions using simply 
%$\vec x(t)=(x(t),y(t))^T$?
% When the parameters are chosen appropriately close
%orbits should emerge that are fixed in time --- in particular the perihelion (the point of closest approach
%of planet and sun) stays fixed in space.
%\end{itemize}
In general an attractive force that vanishes at large distances leads,
 depending on the initial conditions, either to bounded orbits or to open orbits --- in the latter case the planet simply disappears from the sun.
Depending on time available for the course  it might be helpful to explain to the students the origin of this, however, since the most transparent reasoning needs an integration as well as the concept of angular momentum not necessarily familiar to the whole class, it appears in general more appropriate to work out together with the students that a moving body can not be kept by a force if its momentum is too large.
Then they may try to find closed orbits in the simulation, e.g., by varying the start velocity while keeping the start location fixed.

@@

The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
100 earth years~\footnote{The symbol $''$ denotes 'arc seconds': 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of $\delta \phi_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
predictions of General Relativity to this particular observable. \todo{citation here}

We allow for a movement of the perihelion by modifying the potential of Eq.~(\ref{eq:newton}) with the following ansatz
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}\right) \ ,
\label{eq:newton_art}
\end{equation}
where $\alpha$ is some parameter related to the effects by General Relativity.
The idea is that the correction must be dimension-less and should scale with some inverse power of $r$, for we still need to demand that the potential vanishes at large distances.
Since the only parameter of the system with dimension of a length is $r_S$, the correction should scale as $(r_S/r)$.
Replacing $r$ by a typical distance of Mercury to the sun, $(r_S/r_{MS})\sim 10^{-7}$, suggests that it should be sufficient to include this kind of term to first order only.\footnote{Considering another order would lead qualitatively to the same result, since every change in the potential leads to a perihelion motion. Only the value of the pre-factor would change.}
For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
\begin{equation}
	\delta \phi_V \approx \frac{1}{T_V} \delta \phi_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
\end{equation}
where $T_V$ ($T_M$) denotes the length of a venus (mercury) year. This precisely describes the experimental value of $9''$.

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python~\cite{}\todo{Citation} as  programming language, because \python is easy to learn, 
intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction 
how to set up \python on different operating systems.}.  No prior knowledge of \python or any other programming language 
is required, because we explain all the steps necessary to create the simulation.  We also provide a working 
example \cite{}
%
\todo{One can additional files to the arXiv upload} that can be used as template.
%

To start the simulation one needs the "initial" distances and velocities, which are needed as input to the equation of motion (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).  As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s \footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.  Since the computer does not understand physical units, one has to express each variable in an appropriate unit.  Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".  A useful choice for this problem is given e.g. by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one mercury year is given by $T_M=88\,T_0$.  With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become
\todo{Maybe we should rename $a_M$ to avoid confusion about the fact, that it is not an acceleration?}
\begin{equation}
r_{MS}(0) = 4.6 R_0 \, , \quad
v_{M}(0)  = 0.51 \frac{R_0}{T_0} \, ,  \quad
a_M       = 0.99 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2} 
\, .
\end{equation}

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.4\textwidth]{figs/sun_merc.pdf}
	\caption{\label{fig:sun_merc}Sun mercury system with relevant vectors.  Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.}
\end{figure}
%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt, UK english]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Custom
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[T1]{fontenc}           % Embed fonts
\usepackage{textcomp,gensymb}      % \degree

\usepackage{polyglossia}                         % Define languages (Umlaute)
\setdefaultlanguage[variant=american]{english}   % Accent of main language
\setotherlanguage{german}                        % We also have german words
\usepackage[autostyle=true]{csquotes}            % Define quotation style
\MakeOuterQuote{"}                               % Text wrapped in "..." is recognized as a quotation

\usepackage{amssymb}                  % Math stuff
\usepackage{graphicx}                 % Import Pictures + PDFs
\usepackage[svgnames]{xcolor}         % Enabling colors by their 'svgnames'
\usepackage{listings}                 % Code listings  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Definitions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\definecolor{fzjblue}{HTML}{005B81}
\definecolor{light-gray}{gray}{0.97}
\definecolor{semilight-gray}{gray}{0.67}
\definecolor{mygreen}{rgb}{0,0.6,0}

% Custom code listings environment
\lstset{
  language=Python,
  backgroundcolor=\color{light-gray},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
  basicstyle=\linespread{1}\footnotesize\ttfamily,        % the size of the fonts that are used for the code
  commentstyle=\color{mygreen},
  frame=single,
  keywordstyle=\color{blue},       % keyword style
  tabsize=4,	                   % sets default tabsize to 2 spaces
  showtabs=false,
  rulecolor=\color{black},
  columns=fullflexible,
  literate={\ }{{\copyablespace}}1
}

\usepackage[prependcaption, textsize=tiny]{todonotes}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Commands
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\python}[0]{\texttt{Python} }
\newcommand{\abs}[1]{\left\vert #1 \right\vert}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Here it starts
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\author{I. Hammer, C. Hanhart, C. Körber and C. Müller}

\address{Forschungszentrum Jülich}
\ead{c.hanhart@fz-juelich.de}
\vspace{10pt}
%\begin{indented}
%\item[]February 2014
%\end{indented}

@@

\listoftodos

@@

We are convinced that in order to excite the students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
This was already demonstrated in the "Schülderakademie Teilchenphysik", where an early version of this course was tested successfully on a group of 10 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.

@@

We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories observed.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space) the location is parametrized by a single three vector $\vec r(t)$.
As will become clear below, in order to describe and control the dynamics of a physical object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{dt}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced with the last expression a common short hand notation for time derivatives.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law.
Said differently: a force $\vec F$ shows up by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation} 
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as the rocket rises.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation} 
Note that in general the force could depend also on the time or the velocity.
We here restrict ourselves to the case relevant for our example, where the force depends on the location only. 
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g. solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general a differential equation of $n^{\rm th}$ degree (where the highest derivative is $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some defined time, but one may as well pick two location at different times.%
}

For this procedure to work $\Delta t$ must be sufficiently small.
What this means depends on the process studied.
One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\ . 
\label{eq:check}
\end{equation}
holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newtons law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec r(t)|$ denotes the distance between sun and Mercury, when we assume that the sun is infinitely more heavy than the planet and located at the center of the coordinate system. 
Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
This is what we need $r_S$ for later in the discussion.
That the Schwarzschild radius is also an important quantity to characterize black holes is not relevant for the discussion at hand.
With this Newtons second law reads 
\begin{equation}
	\ddot{r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \, , \qquad
	\ddot{\vec r} =   \ddot{r} \, \frac{\vec{r}}{r} \, .
\label{eq:newton}
\end{equation}
In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded orbits or to open orbits --- in the latter case the planet simply disappears from the sun.
Depending on time available for the course it might be helpful to explain to the students the origin of this, however, since the most transparent reasoning needs an integration as well as the concept of angular momentum not necessarily familiar to the whole class, it appears in general more appropriate to work out together with the students that a moving body can not be kept by a force if its momentum is too large.
Then they may try to find closed orbits in the simulation, e.g., by varying the start velocity while keeping the start location fixed.

@@

The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
100 earth years~\footnote{The symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of $\delta \Theta_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
predictions of General Relativity to this particular observable. \todo{@CH:citation here}

We allow for a movement of the perihelion by modifying the potential of Eq.~(\ref{eq:newton}) with the following ansatz
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}\right) \, \frac{\vec{r}}{r} \ ,
\label{eq:newton_art}
\end{equation}
where $\alpha$ is some parameter related to the effects by General Relativity.
The idea is that the correction must be dimension-less and should scale with some inverse power of $r$, for we still need to demand that the potential vanishes at large distances.
Since the only parameter of the system with dimension of a length is $r_S$, the correction should scale as $(r_S/r)$.
Replacing $r$ by a typical distance of Mercury to the sun, $(r_S/r_{MS})\sim 10^{-7}$, suggests that it should be sufficient to include this kind of term to first order only.\footnote{Considering another order would lead qualitatively to the same result, since every change in the potential leads to a perihelion motion. Only the value of the pre-factor would change.}
For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
\begin{equation}
	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
\end{equation}
where $T_V$ ($T_M$) denotes the length of a venus (mercury) year. This precisely describes the experimental value of $9''$.

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{}\footnote{\url{https://www.python.org}} as  programming language, because \python is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python or any other programming language is required, as we explain all the necessary steps to create the simulation.
In appendix \ref{appendix:python} we give instructions for setting up \python on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{url once uploaded (arXiv)}.

To start the simulation one needs the "initial" distances and velocities, which are needed as input to the equation of motion [Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})].
As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.

Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s\footnote{\url{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}}.
Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
A useful choice for this problem is given e.g. by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one mercury year is given by $T_M=88\,T_0$.
With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration pre-factors become
\begin{eqnarray}
r_{MS}(0)   &=& 4.6 R_0 \, , \quad
v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
a_M(r_{MS}) &=& 1.01 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2} 
\, .
\end{eqnarray}

%
\begin{figure}[htb]
	\centering
	\includegraphics[width=.4\textwidth]{figs/sun_merc.pdf}
	\todo[inline]{@CK: figure: coordinate system}
	\caption{\label{fig:sun_merc}%
		Sun mercury system with relevant vectors.
		Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
	}
\end{figure}
%
